%!TEX root = ../thesis.tex
%*******************************************************************************
%****************************** Notation ***************************************
%*******************************************************************************

\section*{Preface}
\label{sec:Preface}

A detailed statement of originality and contribution is provided in Chapter 2 and a summary is given below:

\begin{itemize}
    \item Chapter 1 presents my own introduction to the fields of generative modelling and self-supervised learning.
    
    \item Chapter 2 outlines the structure of the thesis and provides a detailed statement of originality and contribution.
    
    \item Chapter 3 contains original research based on the publication \cite{batzolis2024caflow}, developed in collaboration with Marcelo Carioni, Christian Etmann, Soroosh Afyouni, Zoe Kourtzi and Carola-Bibiane Sch\"onlieb. This work is published in the \textit{Foundations of Data Science (FoDS)} journal, AIMS, 2024.
    
    \item Chapter 4 presents original research based on the article \cite{batzolis2022non_uniform}, conducted in collaboration with Jan Stanczuk, Carola-Bibiane Sch\"onlieb and Christian Etmann. The article is to be submitted for publication.
    
    \item Chapter 5 presents original research based on the publication \cite{pmlr-v235-stanczuk24a}, developed in collaboration with Jan Stanczuk, Teo Deveney and Carola-Bibiane Sch\"onlieb. This work is published in the \textit{Proceedings of the 41st International Conference on Machine Learning (ICML)}, Vienna, Austria. PMLR 235, 2024.
    
    \item Chapter 6 presents original research based on the article \cite{batzolis2023variational}, developed in collaboration with Jan Stanczuk and Carola-Bibiane Sch\"onlieb. The article is to be submitted for publication.
    
    \item Chapter 7 presents original research based on the publication \cite{diepeveen2024score}, developed in collaboration with Willem Diepeveen, Zakhar Shumaylov and Carola-Bibiane Sch\"onlieb. This work is published in the \textit{Proceedings of the 42nd International Conference on Machine Learning (ICML)}, Vancouver, Canada. PMLR 267, 2025.
    
    \item Chapter 8 provides the conclusions and outlook of the thesis.
\end{itemize}

