% ************************** Thesis Acknowledgements **************************

\begin{acknowledgements}      

First and foremost, I would like to express my deepest gratitude to my supervisor, Carola-Bibiane Schönlieb. Your exceptional supervision, guidance, and unwavering support have made this intellectually stimulating and incredibly rewarding journey possible. I am also deeply grateful for your empathy, understanding, and encouragement during challenging times, which have made all the difference. Your guidance and expertise have been a constant source of inspiration, but it is the way you lead your life—with sincere empathy, kindness, integrity, and remarkable wisdom—that truly inspires me on both a professional and personal level.

I would also like to sincerely thank Maria Skoularidou for generously providing valuable feedback that greatly contributed to improving my thesis. Additionally, I gratefully acknowledge the financial support of GSK.

I am deeply grateful to Jan Stanczuk, my primary collaborator and close friend. I will always treasure our endless conversations, which have traversed the realms of science, philosophy, and religion, to the heights of mountaineering, the mysteries of consciousness, and beyond. These discussions have shaped this journey in ways I will never forget. However, it was our shared obsession with diffusion models that truly defined our research and made this journey both fulfilling and productive. I am incredibly proud of what we have accomplished together, and I am truly thankful for the deep friendship that has grown from our professional collaboration.

A special thank you to Lambros Gavalakis for being my closest friend throughout this journey celebrating my happiest moments and standing by me through the most challenging ones. I will always cherish the midday gossip phone calls and the late-night chess games at CMS, our frequent visits to Darwin bar while in Cambridge and our deep, thought-provoking conversations—sometimes filled with spirited debates—over a glass of tsipouro in Athens, Thessaloniki, or Methoni.

A heartfelt thank you to Teo Papazachariou, who has been a true and unwavering friend throughout many years of my life. I am deeply grateful for your support during some of my most difficult times. Sharing the "Raskolnikov" apartment in Cambridge during your postdoc visit was one of the most incredible chapters of my life—memories I will always hold close to my heart.

I would also like to sincerely thank Alexandros Terzopoulos, Dimitrios Lolas, and Ilja Uzelac Bujišić for being such close and supportive friends throughout this journey. Additionally, I am deeply grateful to my dear friends—Panagiotis Kosmas, Renos Lyssiotis, Dimitris Tsitselas, Anastasis Togkousidis, Panos Bogiantzis, Spyros Mantzavinos, Chris Manikis, Nick Manikis, Apostolos Panagiotopoulos, Vicky Parousidou, Giorgos Fotiadis, Zoi Tsangalidou, Yiannis Papageorgiou, Alexandros Tzikas, Andreas Economou and Konstantinos Kavvadias—whose support has meant so much to me, even when circumstances didn’t allow us to meet as often as we would have liked.

I extend my deepest gratitude to my partner, Maria Androudi, who has been a constant source of strength and inspiration for me with her love, support, and unwavering belief in me. Your presence has truly enriched my life, and I am endlessly grateful for you.

I am eternally grateful to my parents, Aspasia and Nikolaos, for their unconditional love and support throughout this journey and generally throughout my life. I could not have completed this journey without you. I also wish to sincerely thank my sister, Maria Batzoli, for always standing by my side and offering her constant support.

I wish to express my heartfelt gratitude to Father Anastasios, my lifelong spiritual father, whose support has been invaluable during the most challenging times of my life. I am deeply thankful for his guidance in helping me find my path to God, whom I believe to be the ultimate source of strength, wisdom, and meaning in this world. Above all, I offer my deepest gratitude to God for His infinite grace, divine guidance, and unwavering presence in my life.

\end{acknowledgements}
